# Visualizations

```{r vis_startup,echo=FALSE,message=FALSE,warning=FALSE}
library(caret)
library(knitr)
opts_chunk$set(digits = 3, tidy = FALSE, fig.path = 'visualizations/', dev = 'svg', dev.args = list(bg = "transparent"))
library(lattice)
library(ggplot2)
```


The `featurePlot` function is a wrapper for different [`lattice`](http://cran.r-project.org/web/packages/lattice/index.html) plots to visualize the data. For example, the following figures show the default plot for continuous outcomes generated using the `featurePlot` function.

For classification data sets, the `iris` data are used for illustration.

```{r iris}
str(iris)
``` 

**Scatterplot Matrix**


```{r VisScatMat,fig.width=6.5,fig.height=6.5,tidy=FALSE, message=FALSE,warning=FALSE}
library(AppliedPredictiveModeling)
transparentTheme(trans = .4)
library(caret)
featurePlot(x = iris[, 1:4], 
            y = iris$Species, 
            plot = "pairs",
            ## Add a key at the top
            auto.key = list(columns = 3))
``` 
    

**Scatterplot Matrix with Ellipses**


```{r VisScatMatEl,fig.width=6.5,fig.height=6.5,tidy=FALSE}
featurePlot(x = iris[, 1:4], 
            y = iris$Species, 
            plot = "ellipse",
            ## Add a key at the top
            auto.key = list(columns = 3))
```

**Overlayed Density Plots**

```{r VisDens,fig.width=9,fig.height=4,tidy=FALSE}
transparentTheme(trans = .9)
featurePlot(x = iris[, 1:4], 
            y = iris$Species,
            plot = "density", 
            ## Pass in options to xyplot() to 
            ## make it prettier
            scales = list(x = list(relation="free"), 
                          y = list(relation="free")), 
            adjust = 1.5, 
            pch = "|", 
            layout = c(4, 1), 
            auto.key = list(columns = 3))
```


**Box Plots**

```{r ViBox,fig.width=9,fig.height=4,tidy=FALSE}
featurePlot(x = iris[, 1:4], 
            y = iris$Species, 
            plot = "box", 
            ## Pass in options to bwplot() 
            scales = list(y = list(relation="free"),
                          x = list(rot = 90)),  
            layout = c(4,1 ), 
            auto.key = list(columns = 2))
```
    
**Scatter Plots**

For regression, the Boston Housing data is used:

```{r bh}
library(mlbench)
data(BostonHousing)
regVar <- c("age", "lstat", "tax")
str(BostonHousing[, regVar])
``` 
    
When the predictors are continuous, `featurePlot` can be used to create scatter plots of each of the predictors with the outcome. For example:

```{r VisScatter,fig.width=9,fig.height=4,tidy=FALSE}
theme1 <- trellis.par.get()
theme1$plot.symbol$col = rgb(.2, .2, .2, .4)
theme1$plot.symbol$pch = 16
theme1$plot.line$col = rgb(1, 0, 0, .7)
theme1$plot.line$lwd <- 2
trellis.par.set(theme1)
featurePlot(x = BostonHousing[, regVar], 
            y = BostonHousing$medv, 
            plot = "scatter", 
            layout = c(3, 1))
```
    
Note that the x-axis scales are different. The function automatically uses `scales = list(y = list(relation = "free"))` so you don't have to add it. We can also pass in options to the [`lattice`](http://cran.r-project.org/web/packages/lattice/index.html) function `xyplot`. For example, we can add a scatter plot smoother by passing in new options:

```{r VisScatterSmooth,fig.width=9,fig.height=4,tidy=FALSE}
featurePlot(x = BostonHousing[, regVar], 
            y = BostonHousing$medv, 
            plot = "scatter",
            type = c("p", "smooth"),
            span = .5,
            layout = c(3, 1))
```
    
The options `degree` and `span` control the smoothness of the smoother.

