```{r sim_load, echo = FALSE, message=FALSE, warning=FALSE, results = 'hide'}
library(caret)
library(proxy)
library(networkD3)
library(heatmaply)
theme_set(theme_bw())
library(knitr)
opts_chunk$set(digits = 3, tidy = FALSE)


modelInfo <- getModelInfo()

tags <- vector(mode = "list", length = length(modelInfo))
Classification <- Regression <- ClassProbs <- VarImpMethod <- rep(NA, length = length(modelInfo))

for(i in seq(along = modelInfo)){
  tags[[i]] <- modelInfo[[i]]$tags
  Classification[i] <- ifelse("Classification" %in% modelInfo[[i]]$type, 1, 0)
  Regression[i] <- ifelse("Regression" %in% modelInfo[[i]]$type, 1, 0)
  ClassProbs[i] <- ifelse(is.null(modelInfo[[i]]$prob), 0, 1)
  VarImpMethod[i] <- ifelse(is.null(modelInfo[[i]]$varImp), 0, 1)
  names(tags)[i] <- paste(modelInfo[[i]]$label, 
                          " (", names(modelInfo)[i],
                          ")", sep = "")
}
tabs <- table(unlist(tags))
tabs <- tabs[order(tolower(names(tabs)))]

terms <- names(tabs)
terms <- terms[terms != ""]
dat <- matrix(0, ncol = length(terms), nrow = length(tags))
colnames(dat) <- terms
hasTag <- lapply(tags, function(x, y) which(y%in%x), y = terms)

for(i in seq(along = hasTag)) {
  dat[i, hasTag[[i]]] <- 1
}
dat <- cbind(Classification, Regression, dat)
rownames(dat) <- names(tags)

write.csv(dat, file = "tag_data.csv")

D <- dist(dat, method = "Jaccard")
Dm <- as.matrix(D)
sim <- 1 - Dm

grps <- rep(NA, nrow(sim))
grps[dat[,"Classification"] == 1 & dat[,"Regression"] == 1] <- 3
grps[dat[,"Classification"] == 0 & dat[,"Regression"] == 1] <- 1
grps[dat[,"Classification"] == 1 & dat[,"Regression"] == 0] <- 2


g_data <- data.frame(source_t = rownames(dat),
                     Group = grps)

id_values <- data.frame(source_t = rownames(dat))
id_values$ID <- as.numeric(id_values$source_t) - 1

###################################################################
##

cutoff <- .7
a <- b <- v <- NULL

for(i in 1:ncol(sim)) {
  tmp <- sim[-i, i]
  if(any(tmp > cutoff)) {
    value <- tmp[tmp > cutoff]
    tmp <- names(tmp)[tmp > cutoff]
    a <- c(a, rep(colnames(sim)[i], length = length(tmp)))
    b <- c(b, tmp)
    v <- c(v, value)
  } else print(colnames(sim)[i])
}

all(unique(a) %in% unique(b))

plot_dat <- data.frame(source_t = a, target_t = b, value = 1)


links <- plot_dat
links$source <- as.numeric(links$source_t) - 1
links$target <- as.numeric(links$target_t) - 1
links <- links[order(links$source, links$target),]

Nodes <- c(as.character(plot_dat$source_t), 
           as.character(plot_dat$target_t))

nodes <- data.frame(source_t = unique(Nodes))
nodes <- merge(nodes, id_values, all.x = TRUE)  
nodes <- merge(nodes, g_data, all.x = TRUE)
names(nodes)[1] <- "name"
nodes <- nodes[order(nodes$ID),]
```

# Models Clustered by Tag Similarity

This page shows a network diagram of all the models that can be accessed by `train`. See the [Revolutions blog](http://blog.revolutionanalytics.com/2014/01/predictive-models-in-r-clustered-by-tag-similarity-1.html) for details about how this visualization was made (and [this page](https://github.com/topepo/caret/blob/master/html/similarity.Rhtml) has updated code using the [`networkD3`](http://cran.r-project.org/web/packages/networkD3/index.html)  package). In summary, the package annotates each model by a set of tags (e.g. "Bagging", "L1 Regularization" etc.). Using this information we can cluster models that are similar to each other. 

Green circles are models only used for regression, blue is classification only and orange is "dual use". Hover over a circle to get the model name and the model code used by the [`caret`](http://cran.r-project.org/web/packages/caret/index.html) package and refreshing the screen will re-configure the layout. You may need to move a node to the left to see the whole name. `r length(modelInfo) - nrow(nodes)` models without connections are not shown in the graph.


```{r makeGraph, echo=FALSE, message=FALSE, screenshot.force = FALSE, results='asis', fig.align="center"}
forceNetwork(Links = links, 
             Nodes = nodes,
             Source = "source",
             Target = "target", 
             Value = "value",
             Group = "Group",
             NodeID = "name",
             charge = -2,
             fontSize = 16,
             height = 600,
             width = 600,
             zoom = TRUE,
             opacity = .8,
             linkDistance = 20
             )
``` 

The data used to create this graph can be found [here](tag_data.csv). 

The plot below shows the similarity matrix. Hover over a cell to see the pair of models and their [Jaccard similarity](https://en.wikipedia.org/wiki/Jaccard_index). Darker colors indicate similar models. 


```{r sim_mat, echo=FALSE, message=FALSE, warning=FALSE, screenshot.force = FALSE, results='asis'}
heatmaply(sim, symm = TRUE, cexRow = .0001, cexCol = .0001, colors = BuPu, branches_lwd = .1)
``` 

You can also use it along with maximum _dissimilarity_ sampling to pick out a diverse set of models. Suppose you would like to use a SVM model with a radial basis function on some regression data. Based on these tags, what other four models would constitute the most diverse set?

```{r dissim}
tag <- read.csv("tag_data.csv", row.names = 1)
tag <- as.matrix(tag)

## Select only models for regression
regModels <- tag[tag[,"Regression"] == 1,]

all <- 1:nrow(regModels)
## Seed the analysis with the SVM model
start <- grep("(svmRadial)", rownames(regModels), fixed = TRUE)
pool <- all[all != start]

## Select 4 model models by maximizing the Jaccard
## dissimilarity between sets of models
nextMods <- maxDissim(regModels[start,,drop = FALSE], 
                      regModels[pool, ], 
                      method = "Jaccard",
                      n = 4)

rownames(regModels)[c(start, nextMods)]
```
